---
title: "data.table package"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# data.table notes
```{r}
library(data.table)

#dataset
msleepDT <- as.data.table(ggplot2::msleep)
irisDT <- as.data.table(iris)
```



## General syntax

`DT[i, j, by]` - take data.table DT, subset rows i, manipulate columns j by something.


## Importing data


## Selecting rows

+ `msleepDT[1:3, ]`: selecting rows 1-3
+ `msleepDT[1:3]`: selecting rows 1-3
+ `msleepDT[.N]`: selecting the last row. `.N` is a shortcut for `nrow()`


Logical subsetting:
+ `msleepDT[sleep_total > 15]`: selecting rows with a column value above 15
+ `msleepDT[vore == "omni"]`: selecting rows with omni vores
+ `msleepDT[vore %in% c("omni", "carni")]`: selecting rows with omni vores

To get to the boolean version: `msleepDT[, sleep_total > 15]`





<br>

## Selecting columns:

+ `msleepDT[,1 ]`: selecting first column in a data.table
+ `msleepDT[, .(genus, vore)]`: selecting genus and vore columns. `.()` is a shortcut for list
+ `msleepDT[, name]`: returns a vector with all names
+ `msleepDT[, .(name)]`: returns a one column data.table with all names



<br>

## Calculations in columns

+ `msleepDT[, mean(sleep_total)]`: calculating the mean, no column name
+ `msleepDT[, .(mean=mean(sleep_total))]`: calculating the mean and giving a colname
+ `msleepDT[, .(mean=mean(sleep_total), median=median(sleep_total))]`: calculating multiple columns
+ `msleepDT[, .(name, mean=mean(sleep_total), median=median(sleep_total))]`: keeping a column, adding the overall mean and median (value gets recylced)


<br>

## Calculations by groups

+ `msleepDT[, .(mean=mean(sleep_total)), by=vore]`: calculating the mean by vore.
+ `msleepDT[, .(mean=mean(sleep_total)), by=.(genus,vore)]`: calculating the mean by vore.


<br>

## Adding and removing columns

These operations update by reference which means your source table is affected and not limited to these lines of code!


+ `msleepDT[, colZ := "Z"]`: adding a column with values Z
+ `msleepDT[, colZ := NULL]`: removing column colZ
+ `msleepDT[, c("awake", "awake_s") := .(awake*60, awake*3600)]`: updating a column and adding a second one


Functional `:=` operator allows to chain operations without specifyin the left hand side upfront.

```{r}
msleepDT[, `:=`(colZ = "Z")(colK = sleep_total/60)(colX = "X")]
msleepDT
```



<br>

## Chaining operations

You can chain multiple operations behind each other by bracketing.

```{r}
msleepDT[, .(mean=mean(sleep_total)), by=vore][order(vore)]
```


<br> 

## Looping through means

If you want to get the mean of multiple columns, there is a more efficient way than to repeat a few times `mean(col)`.

+ `.SD`: shortcut that contains all columns, except the ones mentioned in `by`.
+ `.SDcols`: specifies which columns of DT to include in .SD 

```{r}
irisDT[, lapply(.SD, median), by=Species]
```

Using .SDcols to specify which ones:

```{r}
irisDT[, lapply(.SD, median), .SDcols = c("Sepal.Width","Sepal.Length"), by=Species]
```



<br>

## Changing and reordering columns

+ `setnames(DT, "oldcol", "newcol")`: changing a column name
+ `setnames(DT, tolower(names(DT)))`: all columns to lower case

+ `setcolorder(DT, c("col2", "col1")): swaps ordering
